Degeneracy in a restricted Boltzmann machine

### 2016-02-18 writing/degeneracy/draft.Rmd
added kaiser method on the grid and some writing

#Degeneracy in Random Graph Models
In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. We consider three related definitions of degeneracy as they relate to a graphical model. Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between marginal and conditional expectations of the model.

@@

##Conditional Expectation

@@

@@

In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying magnitudes. The results of this numerical study can be seen in figure \ref{fig:split_plots}. From this figure, it is evident that the fraction of near-degenerate models simulated at specified magnitudes of $\boldsymbol theta_{main}$ and $\boldsymbol theta_{interaction}$ is more closely related to the combined distance from the origin of these two vectors, than any individual group of parameter values.

@@

```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model. For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=3}
plot_dat_grid %>% 
  mutate(dist = sqrt(r1^2 + r2^2)) %>%
  group_by(dist, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n()) %>%
  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
  ggplot() +
  geom_point(aes(dist, frac_degen, colour = Hiddens)) +
  geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
  facet_grid(~Visibles) +
  ylim(c(0,1)) +
  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
  ylab("Fraction Near-degenerate")
```

>>>>

#Degeneracy in Random Graph Models
In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. We consider three related definitions of degeneracy as they relate to a graphical model. Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between expectations of the model and expectations under independence.

@@

@@

##Expectation Differences
The characterization put forth by @kaiser2006markov of degeneracy in a model concerns the difference between model expectations and expectations given independence. Using this concept of dependence in a model, it is possible to understand what conditions lead to near-degeneracy in a model versus those that guarantee non-degenerate models. In this method, a model is considered without near-degeneracy if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

\begin{align*}
\text{E}\left[\boldsymbol X\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}

@@

In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying magnitudes. The results of this numerical study can be seen in figure \ref{fig:split_plots}. From this figure, it is evident that the fraction of near-degenerate models simulated at specified magnitudes of $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ is more closely related to the combined distance from the origin of these two vectors, than any individual group of parameter values.

@@

```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model. For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=2}
plot_dat_grid %>% 
  mutate(dist = sqrt(r1^2 + r2^2)) %>%
  group_by(dist, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n()) %>%
  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
  ggplot() +
  geom_point(aes(dist, frac_degen, colour = Hiddens)) +
  geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
  facet_grid(~Visibles) +
  ylim(c(0,1)) +
  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
  ylab("Fraction Near-degenerate")
```
